A Large-scale Study of Polymorphism in R

### 2019-04-01 notebooks/paper_L123_results_section.Rmd
more prose

3. Extras. What about errors? How often do we have metaprogramming.
```{r}
meta_los <- lapply(los, function(lof) lapply(lof, annihilate_errors_df))
```

>>>>

3. Extra. What about errors? How often do we have metaprogramming.
```{r}
# This can change depending on how we address metaprogramming.
meta_los <- lapply(los, function(lof) lapply(lof, annihilate_errors_df))
```

Step 4. Translate to L3?
What is L3?
We want class info and attribute info, on some level.
Let's do classes first.
```{r}
# get rid of classes we have dealt with.
los <- lapply(los, function(lof) lapply(lof, class_to_primitive_df))
```

How do we want to deal with the classes?
Should they become a type, if they are not primitive?
Let's look at class poly counts maybe.
```{r}
L3_c_basic_df <- build_arg_count_df_from_lopkgs(los)

L3_poly_class_args <- reduce_extracted_lopkgs_to_one(apply_extract_to_lopkg(los, extract_poly_class_args))
L3_poly_class_args <- L3_poly_class_args[order(-L3_poly_class_args$count),] # sort by count
L3_poly_class_args$perc <- round(L3_poly_class_args$count/sum(L3_poly_class_args$count) * 100, 2)
```

All the numbers are low, so there's no clear pattern, which makes sense.
How best to incorporate this...
```{r}

### 2019-04-01 paper/main.tex
more prose

\lstdefinestyle{Rin}{ %
  style=R,
  numberstyle=none,
  basicstyle=\normalsize\ttfamily,
  breaklines=false
}
\newcommand{\code}[1]{\lstinline|#1|\xspace}

\newcommand{\genthat}{{\sc Genthat}\xspace}

@@

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 22 18:03:30 2019
\begin{table}[ht]
\label{tab:argcountsL1}
\centering
\begin{tabular}{lrr}
  \hline
 Argument Polymorphism & Count & Percentage \\ 
  \hline
  Full Monomorphic & 781K & 85.37 \\ 
  Monomorphic in Type & 834K & 91.23 \\ 
  Monomorphic in Class & 827K & 90.42 \\ 
  Monomorphic in Attribute Pattern & 825K & 90.27 \\ 
  Total Seen & 914K & --- \\ 
     \hline
\end{tabular}
\caption{Account of {\it argument} polymorphism in L1.}
\end{table}

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 22 18:10:46 2019
\begin{table}[ht]
\label{tab:funcountsL1}
\centering
\begin{tabular}{lrr}
  \hline
 Function Polymorphism & Count & Percentage \\ 
  \hline
  Full Monomorphic & 143K & 65.88 \\ 
  Monomorphic in Type & 167K & 77.25 \\ 
  Monomorphic in Class & 163K & 75.24 \\ 
  Monomorphic in Attribute Pattern & 161K & 74.37 \\ 
  Total Seen & 216K & --- \\ 
   \hline
\end{tabular}
\caption{Account of {\it function} polymorphism in L1.}
\end{table}

@@

We see that roughly 9\% of all argument positions in R are polymorphic in type.
The numbers for attribute and class polymorphism are included for completeness, but the important number is that of type polymorphism (as our method of constructing types sometimes factors in class and attribute).
Specifically for classes, often the class and \code{typeof} a value will line up (refer to Section~\ref{sec:rlang}).

As for functions, roughly 23\% are polymorphic in type, or put differently 23\% have at least one polymorphic argument or a polymorphic return.
The discrepancy between this and the polymorphicity of arguments is explained by the tendency of some functions to have {\it many} arguments.

These numbers alone are insufficient, so we ask: {\it of the polymorphic arguments, what are the most common patterns?}
The answer to that question can be found in Table~\ref{tab:toppolyL1}.

@@

\AT{This should probably be discussed earlier.}
Not pictured here, the most common polymorphic pattern in L1 is the {\tt scalar/X}, {\tt vector/X} pattern, but this is dealt with separately:
If a function argument is observed to be both a unit-length and non-unit-length vector, then it's reasonable to assume that the argument is intended to be a vector, so we collapse the distinction in such cases.

In Table~\ref{tab:toppolyL1}, we again see the prominence of {\tt double} and {\tt integer} polymorphism.
As discussed previously, we will address this by creating a new {\it real} type encompassing both.
Refer back to the snippet in Figure~\ref{fig:realex} for details.

As we make the distinction between vectors and scalars, one other interesting question that arises is: {\it how many arguments are strictly scalar?}
As it happens, over 40\% of monomorphic arguments with primitive vectorized types are always scalar (so, 40\% of {\it monomorphic} occurrences of {\tt scalar/X} or {\tt vector/X} are scalar).
This shows that there is value in a scalar annotation, which could be used to communicate to an R runtime that vectorization of a value is not necessary, which may in turn lead to a performance improvement.

Another clear pattern in Table~\ref{tab:toppolyL1} is that of NULL, X.
As we discussed, we will roll NULL (and raw NAs) into other types.

The next type system design we will consider is ... .
The number of monomorphic arguments and functions can be found in Tables~\ref{tab:argcountsWHATTS} and~\ref{tab:funcountsWHATTS} respectively.

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 22 18:03:30 2019
\begin{table}[ht]
\label{tab:argcountsWHATTS}
\centering
\begin{tabular}{lrr}
  \hline
 Argument Polymorphism & Count & Percentage \\ 
  \hline
  Full Monomorphic & 781K & 85.47 \\
  Monomorphic in Type & 867K & 95.01 \\
%  Monomorphic in Class & 831K & 90.31 \\ 
%  Monomorphic in Attribute Pattern & 830K & 90.25 \\ 
  Total Seen & 914K & --- \\ 
     \hline
\end{tabular}
\caption{Account of {\it argument} polymorphism in the NEW TYPE SYSTEM.}
\end{table}

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 22 18:10:46 2019
\begin{table}[ht]
\label{tab:funcountsWHATTS}
\centering
\begin{tabular}{lrr}
  \hline
 Function Polymorphism & Count & Percentage \\ 
  \hline
  Full Monomorphic & 143K & 66.04 \\ 
  Monomorphic in Type & 184K & 85.30 \\
%  Monomorphic in Class & 163K & 75.13 \\ 
%  Monomorphic in Attribute Pattern & 161K & 74.31 \\ 
  Total Seen & 216K & --- \\ 
   \hline
\end{tabular}
\caption{Account of {\it function} polymorphism in the NEW TYPE SYSTEM.}
\end{table}

We had some big wins here, with only ~5\% of argument polymorphism unaccounted for.
To get an idea of what more we can do, let's see what the most frequent remaining polymorphic signatures are.
See Table~\ref{tab:toppolyWHATTS}.

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Tue Mar 26 17:46:45 2019
\begin{table}[ht]
\label{tab:toppolyWHATTS}
\centering
\begin{tabular}{lrr}
  \hline
Polymorphic Type Signature & \# Occurrences & \% of Total Polymorphism \\ 
  \hline
  matrix, vector/real & 6360 & 13.94 \\ 
  list$<$any$>$, list$<$real$>$ & 2799 & 6.13 \\ 
  vector/character, vector/real & 2774 & 6.08 \\ 
  scalar/character, scalar/real & 2552 & 5.59 \\ 
  data.frame, matrix & 2167 & 4.75 \\ 
  list$<$any$>$, list$<$list$>$ & 1496 & 3.28 \\ 
  list$<$any$>$, list$<$character$>$ & 1214 & 2.66 \\ 
  data.frame, vector/real & 1046 & 2.29 \\ 
  scalar/logical, scalar/real & 920 & 2.02 \\ 
  scalar/character, vector/real & 868 & 1.90 \\
   \hline
\end{tabular}
\caption{Top 10 argument signatures in the NEW TYPE SYSTEM.}
\end{table}

@@

@@

>>>>

\lstdefinestyle{Rin}{ %
  style=R,
  numberstyle=none,
  basicstyle=\normalsize\ttfamily,
  breaklines=false
}

\newcommand{\code}[1]{\lstinline|#1|\xspace}
\newcommand{\genthat}{{\sc Genthat}\xspace}

@@

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 22 18:03:30 2019
\begin{table}[ht]
\label{tab:argcountsL1}
\centering
\begin{tabular}{lrr}
  \hline
 Argument Polymorphism & Count & Percentage \\ 
  \hline
  Full Monomorphic & 781K & 85.37 \\ 
  Monomorphic in Type & 834K & 91.23 \\ 
%  Monomorphic in Class & 827K & 90.42 \\ 
%  Monomorphic in Attribute Pattern & 825K & 90.27 \\ 
  Total Seen & 914K & --- \\ 
     \hline
\end{tabular}
\caption{Account of {\it argument} polymorphism in L1.}
\end{table}

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 22 18:10:46 2019
\begin{table}[ht]
\label{tab:funcountsL1}
\centering
\begin{tabular}{lrr}
  \hline
 Function Polymorphism & Count & Percentage \\ 
  \hline
  Full Monomorphic & 143K & 65.88 \\ 
  Monomorphic in Type & 167K & 77.25 \\ 
%  Monomorphic in Class & 163K & 75.24 \\ 
%  Monomorphic in Attribute Pattern & 161K & 74.37 \\ 
  Total Seen & 216K & --- \\ 
   \hline
\end{tabular}
\caption{Account of {\it function} polymorphism in L1.}
\end{table}

@@

We see that roughly 9\% of all argument positions in L1 are polymorphic in type, as opposed to 7.5\% in L0.
This small difference is perhaps surprising, as we did define a number of additional types and might expect for there to be a lot more polymorphism.
This is explained by the fact that drawing the distinctions added a lot of polymorphism to already polymorphic arguments.
\AT{We can get numbers on this.}

As for functions, roughly 23\% are polymorphic in type, or put differently 23\% have at least one polymorphic argument or a polymorphic return.
Again, the small difference between L0 and L1 in this regard is explained by a tendency to make already polymorphic functions more polymorphic.

As before, these numbers alone are insufficient, so we ask again: {\it of the polymorphic arguments, what are the most common patterns?}
The answer to that question can be found in Table~\ref{tab:toppolyL1}.

@@

We would like to make one important note before digesting the data in Table~\ref{tab:toppolyL1}.
We drew a distinction between scalars and vectors of the same type, and that distinction would be the single largest source of polymorphism in L1 had we not dealt with it appropriately.
We note that an argument which exhibits a signature including e.g. \{scalar/double, vector/double\} really must be an argument designed to take vectors, as recall that unit length vectors appear as scalars by our definition.
We have already collapsed this distinction.

In Table~\ref{tab:toppolyL1}, we again see the prominence of {\tt double} and {\tt integer} polymorphism.
As discussed previously, we will address this by creating a new {\it real} type encompassing both.
Refer back to the snippet in Figure~\ref{fig:realex} for details.
Also, another clear pattern in Table~\ref{tab:toppolyL1} is that of NULL, X.
As we discussed, we will roll NULL (and raw NAs) into other types.

%
%
\subsubsection{Scalars}

As we make the distinction between vectors and scalars, one other interesting question that arises is: {\it how many arguments are strictly scalar?}
As it happens, over 40\% of monomorphic arguments with primitive vectorized types are always scalar (so, 40\% of {\it monomorphic} occurrences of {\tt scalar/X} and {\tt vector/X} are scalar).
This shows that there is value in a scalar annotation, which could be used to communicate to an R runtime that vectorization of a value is not necessary, which may in turn lead to a performance improvement.

At this point, we have seen a number of reasonable polymorphic patterns that arise in R, both in L0 and L1.
The next type system design we consider will deal with these, and we refer to it as L$1^{+}$.
The number of monomorphic arguments and functions can be found in Tables~\ref{tab:argcountsL1p} and~\ref{tab:funcountsL1p} respectively.

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 22 18:03:30 2019
\begin{table}[ht]
\label{tab:argcountsL1p}
\centering
\begin{tabular}{lrr}
  \hline
 Argument Polymorphism & Count & Percentage \\ 
  \hline
  Full Monomorphic & 781K & 85.47 \\
  Monomorphic in Type & 867K & 95.01 \\
%  Monomorphic in Class & 831K & 90.31 \\ 
%  Monomorphic in Attribute Pattern & 830K & 90.25 \\ 
  Total Seen & 914K & --- \\ 
     \hline
\end{tabular}
\caption{Account of {\it argument} polymorphism in L$1^{+}$.}
\end{table}

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 22 18:10:46 2019
\begin{table}[ht]
\label{tab:funcountsL1p}
\centering
\begin{tabular}{lrr}
  \hline
 Function Polymorphism & Count & Percentage \\ 
  \hline
  Full Monomorphic & 143K & 66.04 \\ 
  Monomorphic in Type & 184K & 85.30 \\
%  Monomorphic in Class & 163K & 75.13 \\ 
%  Monomorphic in Attribute Pattern & 161K & 74.31 \\ 
  Total Seen & 216K & --- \\ 
   \hline
\end{tabular}
\caption{Account of {\it function} polymorphism in L$1^{+}$.}
\end{table}

We had some big wins here, with only ~5\% of argument polymorphism unaccounted for.
To get an idea of what more we can do, let's see what the most frequent remaining polymorphic signatures are.
See Table~\ref{tab:toppolyL1p}.

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Tue Mar 26 17:46:45 2019
\begin{table}[ht]
\label{tab:toppolyL1p}
\centering
\begin{tabular}{lrr}
  \hline
Polymorphic Type Signature & \# Occurrences & \% of Total Polymorphism \\ 
  \hline
  matrix, vector/real & 6360 & 13.94 \\ 
  list$<$any$>$, list$<$real$>$ & 2799 & 6.13 \\ 
  vector/character, vector/real & 2774 & 6.08 \\ 
  scalar/character, scalar/real & 2552 & 5.59 \\ 
  data.frame, matrix & 2167 & 4.75 \\ 
  list$<$any$>$, list$<$list$>$ & 1496 & 3.28 \\ 
  list$<$any$>$, list$<$character$>$ & 1214 & 2.66 \\ 
  data.frame, vector/real & 1046 & 2.29 \\ 
  scalar/logical, scalar/real & 920 & 2.02 \\ 
  scalar/character, vector/real & 868 & 1.90 \\
   \hline
\end{tabular}
\caption{Top 10 argument signatures in L$1^{+}$.}
\end{table}

@@

\AT{list<list> might be objects?}

@@

We have seen that once we have done away with trivial sources of polymorphism, a number of interesting patterns arise.
In many of these cases, we can appeal to other sources of reflection in R to expand our understanding.
For example, to deal with the possibility of dates, we can appeal to the \code{class} of values, and to gain more information on matrices we might be interested in looking at attributes.
The next type system addresses these concerns.

%
%
%
%
\subsection{L2}

\AT{Eyy time for L2.}
